A Method for the Spectroscopic Inference of Stellar Parameters

### 2014-07-21 tex/paper.tex
updated sect 2, some notation

\newcommand{\fundamentalParameters}{\vt_{\ast}}
\newcommand{\gridParameters}{\{\fundamentalParameters\}^{\rm 
grid}}
\newcommand{\observationalParameters}{\vt_{\textrm{obs}}}
\newcommand{\stellarParameters}{\vt}

%\phi and \Phi are used for nuisance parameters
\newcommand{\cheb}{ \vp_{\textrm{Cheb}}}
\newcommand{\chebi}[1]{ \vp_{\textrm{Cheb}_{#1}}} %for signifying a specific order
\newcommand{\Cheb}{ \vP_{\textrm{Cheb}}}
\newcommand{\Chebi}[1]{ \vP_{\textrm{Cheb}_{\ne #1}}} %for specifying everything but a specific order
\newcommand{\cov}{ \vp_{\textrm{cov}}}
\newcommand{\covi}[1]{ \vp_{\textrm{cov}_{#1}}} %for signifying a specific order
\newcommand{\Cov}{ \vP_{\textrm{cov}}}
\newcommand{\Covi}[1]{ \vP_{\textrm{cov}_{\ne #1}}} %for specifying everything but a specific order

@@

%kernel notation
\newcommand{\KK}{\mathcal{K}}
\newcommand{\Kglobal}{\KK_{\textrm{g}}}
\newcommand{\Klocal}{\KK_k}

@@

All astronomers recognize that spectroscopy offers a wealth of information that can help
characterize the properties of the observing target.  In the context of stellar astrophysics, 
spectroscopy plays many fundamental roles.  The relative strengths and widths of stellar absorption 
lines provide access to key parameters like effective temperature ($T_{\rm eff}$) and surface 
gravity ($\log g$), enabling model comparisons in the Hertzsprung-Russell diagram to estimate the 
masses and ages so crucial to understanding stellar evolution, as well as individual elemental (and 
molecular) abundances or the collective ``metallicity" (typically parameterized as [Fe/H]), 
facilitating study of the chemical hallmarks of different stellar populations.  With sufficient 
spectral resolution, the velocity content of a spectrum can convey crucial information about 
stellar rotation ($v \sin i$) and kinematics (e.g., association with a cluster or companion through 
the radial velocity, $v_z$).  While many fields benefit from the spectroscopic measurements of 
these stellar properties, there is acute interest in the exoplanet community.  There, all estimates 
of the planet properties are made {\it relative} to the host properties (e.g., the planet-to-host 
mass or radius {\it ratio} is measured with the radial velocity or transit techniques, 
respectively).  Moreover, essential clues to the planet formation process are encapsulated in the 
dependences of planet frequency on host mass \citep[e.g.,][]{johnson07,howard10} and metallicity 
\citep[e.g.,][]{fischer05,buchhave14}.   
% SA: we ought to check with exoplanet people on these references. Can do later...

@@

The remainder of this section describes the mechanics of this modeling framework.  First, a model 
spectrum is generated for a given set of physical parameters (Section \ref{subsec:synthetic}), and 
then post-processed to mimic reality using a set of observational and practical nuisance parameters 
(Section \ref{subsec:postprocess}).  Next, a direct, pixel-by-pixel comparison between the data and 
model spectra is made with a prescribed likelihood function and a parametric treatment of the 
covariances between pixel residuals (Section \ref{subsec:likelihood}).  That process is iterated 
under the guise of hierarchical Monte Carlo Markov Chain (MCMC) simulations to numerically explore 
the posterior probability density of the model conditioned on the data, and thereby to determine 
constraints on the parameters of interest (Section \ref{subsec:MCMC}).  Along the way, these 
procedures are illustrated with real observations of the high resolution optical spectrum from a 
nearby F star.  That specific application, along with some alternative demonstrations of the 
method, are discussed in more detail in Section \ref{sec:examples}.

@@

There are various approaches to synthesizing a spectrum, $f_{\lambda}$, for a specific set of model
parameters, $\vt_{\ast}$.  In an ideal case, a model stellar atmosphere is constructed and then 
subsequently processed through a radiative transfer code \citep[e.g.,][]{kurucz93,hauschildt99}.  
However, in general this approach is still computationally prohibitive for any iterative method of 
probabilistic inference.  One partial compromise is to interpolate over a library of atmosphere 
structures that were pre-computed for a discrete grid of parameter values, $\gridParameters$, for some arbitrary $\fundamentalParameters$, and then perform a radiative transfer calculation with 
that interpolated atmosphere to synthesize $\flam$ \citep[e.g., as for {\tt 
SME};][]{valenti96}.  A more common variant is to instead rely on interpolation over a library of 
pre-synthesized model spectra, $\flam(\gridParameters)$ \citep[e.g.,][]{castelli04,
allard12,husser13}.  While technically the former approach is most similar to the ideal case, the 
computational cost of repeated spectral synthesis is sufficiently high to make a detailed 
exploration of parameter space (particularly for data with a large spectral range) considerably 
less appealing.  A related, but different approach is to eschew forward modeling entirely (and 
therefore repeated spectral syntheses and/or library interpolations), and instead evaluate the 
models only at the discrete grid points of the library.  Then, these discretized samples of the 
posterior probability density can be interpolated to an arbitrary $\fundamentalParameters$ to construct 
appropriate confidence intervals \citep[similar to the method of {\tt SPC};][]{buchhave12}.  The 
difficulty with this latter approach is that the parameter uncertainties can be smaller than the 
grid spacing; in that case, there is valid concern that this interpolation might not accurately 
recover intrinsic parameter degeneracies.

Here we opt to take the computationally expedient approach that employs a library of model spectra, 
$\flam(\gridParameters)$, where $\vt_{\ast} = [T_{\rm eff}, \,\, \log{g}, \,\, Z]$ 
(in practice, the metallicity $Z$ is often parameterized by [Fe/H]).  However, it is worth noting 
that the techniques we will develop are applicable to {\it any} reasonable ``front-end" that 
generates a model spectrum.  In our adopted approach, the model spectrum for an arbitrary 
$\vt_{\ast}$ must be interpolated from among the spectral library, 
\begin{equation}
\flam(\gridParameters) \leadsto \flam(\fundamentalParameters),
\label{eqn:interp}
\end{equation}
where we assign the symbol $\leadsto$ as an interpolation operator.  The multi-dimensional 
interpolation in Eq.~\ref{eqn:interp} needs to be performed many times, so computational efficiency 
is critical.  In practice, a simple tri-linear interpolation is suitably fast, but introduces an 
undesireable level of inaccuracy (particularly in the $Z$ dimension).  The interpolation quality 
can be empirically estimated by performing the operation in Eq.~\ref{eqn:interp} across a 
calculated location in $\{\vt_{\ast}\}^{\rm grid}$, and then comparing the interpolated spectrum 
with the corresponding library spectrum.  After an extensive exploration of such calculations 
\citep[see also][]{husser12}, we concluded that the best combination of speed and accuracy can be 
achieved by pre-computing a {\it refined} spectral library using a cubic spline interpolation with 
a $\{\vt_{\ast}\}^{\rm grid}$ spacing of [20\,K, 0.1\,dex, 0.1\,dex], and then performing 
tri-linear interpolation over that refined grid.  Overall, this interpolation technique is found to 
be accurate within a few percent per high resolution model pixel.  Ideally, this pre-interpolation 
could be avoided if the spectral library was computed over a refined grid (with a substantial 
up-front computational investment); but for the time being, we can empirically propagate these 
interpolation uncertainties into the likelihood calculations (as will be described in 
Section~\ref{subsec:likelihood}).  
\comm{I see nothing was done here regarding your updated approach; how should we handle that?}

@@

Generally, the ``raw" model spectrum $\flam(\fundamentalParameters)$ will be highly over-sampled compared 
to a typical observed spectrum, and does not account for several additional observational and 
instrumental effects that become important in comparisons with real data.  Therefore, a certain 
amount of post-processing is required before assessing the model quality.  We treat that 
post-processing in two stages: the first deals with an additional set of ``observational" 
parameters, $\observationalParameters$, that incorporate dynamical effects, geometry, and the relative 
location of the target, while the second employs a suite of nuisance (hyper-)parameters, $\Cheb$, designed to mitigate an imperfect data calibration.

We can further divide $\observationalParameters$ into those parameters that impact the model primarily in the 
spectral or flux dimensions; $\observationalParameters = [\vt_{\rm obs, v}, \,\, \vt_{\rm obs, f}]$.  For the 
former, we consider three kernels that contribute to the observed line-of-sight velocity 
distribution function, $\varphi_v$.  The first, $\mathcal{F}_v^{\rm inst}$, treats the instrumental 
spectral broadening.  For illustrative purposes we assume $\mathcal{F}_v^{\rm inst}$ is a Gaussian 
with a mean at $v = 0$ and a constant width $\sigma_v$ at all $\lambda$, although more 
sophisticated forms could be adopted.  The second, $\mathcal{F}_v^{\rm rot}$, characterizes the 
broadening induced by (projected) stellar rotation, parameterized by $v\sin{i}$ as described by 
\citet[][his Eq.~18.14]{gray08}.  And the third, $\mathcal{F}_v^{\rm dop} = \delta(v-v_z)$, 
incorporates the radial velocity through a Doppler shift.  The model spectrum is modified by the 
parameters $\vt_{{\rm obs}, v} = [\sigma_v, \,\, v\sin{i}, \,\, v_z]$ through these kernels, using 
a convolution in velocity-space,\footnote{In practice, these convolutions are performed as 
multiplications in Fourier-space to better preserve spectral information \citep[cf.,][]{tonry79}; 
the mathematical formalism is presented here for clarity.}
\begin{equation}
\flam(\fundamentalParameters, \vt_{{\rm obs}, v}) = \flam(\fundamentalParameters) \otimes \varphi_v = \flam(\fundamentalParameters) \otimes \mathcal{F}_v^{\rm inst} \otimes \mathcal{F}_v^{\rm rot} \otimes \mathcal{F}_v^{\rm dop},
\label{eqn:broadening}
\end{equation}
and then re-sampled onto the discrete wavelengths corresponding to each data pixel, 
\begin{equation}
\flam(\fundamentalParameters, \vt_{{\rm obs}, v}) \mapsto \vM(\fundamentalParameters, \vt_{{\rm obs}, v}),
\label{eqn:resampling}
\end{equation}
where the $\mapsto$ symbol denotes a re-sampling operator that maps the model spectrum onto the 
$N_{\rm pix}$-dimensional array $\vM$ (and $N_{\rm pix}$ is the number of pixels in the spectrum).  
For reference, Figure \ref{fig:broadening} shows a graphical representation of these 
post-processing steps.

@@

At this stage, the model is further modified in the flux dimension.  A typical synthetic spectrum 
is computed as the flux that would be measured {\it at the stellar surface}, and so needs to be 
diluted by the subtended solid angle, $\Omega = (R_{\ast}/d)^2$, where $R_{\ast}$ is the stellar 
radius and $d$ is the distance.  An additional wavelength-dependent scaling factor is applied to 
account for interstellar extinction, assuming some previously-derived extinction law $A_{\lambda}$ 
\citep[e.g.,][]{cardelli89} that is parameterized by $A_V$.  The parameters $\vt_{{\rm obs}, f} = 
[\Omega, \,\, A_V]$ are applied as
\begin{equation}
\vM(\vt) \equiv \vM(\vt_{\ast}, \vt_{\rm obs}) = \vM(\vt_{\ast}, \vt_{{\rm obs}, v}) \times \Omega \times 10^{-0.4\,A_{\lambda}},
\label{eqn:scaling}
\end{equation}
where we have simplified the notation by writing $\vt \equiv [\vt_{\ast}, \,\, \vt_{\rm obs}]$.

So far, the procedure summarized in Eq.~\ref{eqn:interp}-\ref{eqn:scaling} is composed of 
relatively straightforward operations demanded by practical astronomical and computing issues.  If 
the observed spectrum was {\it perfectly} calibrated, we could proceed to a likelihood calculation 
(Section~\ref{subsec:likelihood}) that makes a direct comparison with $\vM(\vt)$ at this point.  
However, that is unlikely to be the case.  The primary concern is that an imperfect calibration 
produces (presumably low-level) mismatches in the underlying shape of the observed spectrum on 
relatively broad wavelength scales.  When compared with the model, these mismatches might represent 
a non-trivial contribution to the likelihood, and thereby bias our estimates of the desired 
physical parameters.  Often, this concern is treated externally to any modeling procedure, usually 
by dividing the observed spectrum (and model spectrum) by an appropriate polynomial function.  But 
that ``normalization" procedure implicitly assumes that there is no relevant information content on 
those mismatched scales; if $\vt$ also contributes to the broad spectral shape, then adopting this 
approach will corrupt the inferences of these parameters.  Moreover, in practice this approach is 
limited, since defining an appropriate polynomial becomes difficult in cases where the spectral 
line density is high (e.g., molecular bands for cool stars).

We adopt a somewhat analogous approach to deal with this issue, but cast it {\it internal} to the 
modeling framework to appropriately propagate the uncertainty introduced by additional degrees of 
freedom in the model.  The residual calibration errors are treated as an explicit contribution to 
the model spectrum, enabling us to explore the distribution of possible calibrations (see 
Section~\ref{subsec:MCMC}) and then eventually marginalize out the associated nuisance parameters.  
In essence, the inferences for the relevant stellar parameters will properly account for the 
underlying uncertainty in the calibration process.  For a continuous spectrum,
such as a long-slit spectrum or a single spectral order of an echelle spectrum,
this is achieved in practice by distorting the model spectrum with a
(low-degree) Chebyshev polynomial \citep[e.g.,][]{eisenstein06,koleva09}, 
\begin{equation}
\vM(\stellarParameters, \cheb) = \vM(\stellarParameters) \times \sum_n c_n \, T_n,
\label{eqn:chebyshev}
\end{equation}
where $T_n$ are the standard Chebyshev functions of degree $n$ and the coefficients are treated as a
set of nuisance (hyper-)parameters, $\cheb = \{c_0, c_1, \ldots, c_{n-1} \}$. For an echelle spectrum with $N$ spectral orders, there is a Chebsyhev polynomial for each order of the spectrum $i$, such that the total collection of Chebyshev parameters is $\Cheb = \{\chebi{1}, \chebi{2}, \ldots, \chebi{N} \}$.  With judicious priors on $\Cheb$, we can ensure that the 
unintended treatment of real spectral features (e.g., broad, deep molecular bands) as calibration 
artifacts is negligible (see Section~\ref{sec:examples} for examples).  The lowest-degree (scaling) 
coefficient, $c_0$, is by its nature degenerate with the solid angle parameter, $\Omega$.  
Therefore, we enforce an additional constraint by requiring that the mean of the polynomial is 
unity.  For data with a single spectral order, this means setting $c_0 = 1$.  If the goal is to 
model multiple spectral orders, we assign $c_0 = 1$ in an arbitrary order as an anchor, but permit 
the $c_0$ values in other orders to be different (as necessary).  It is worth noting that this 
formalism can, in principle, be extended to develop models for completely uncalibrated spectra, 
rather than the residual calibration mismatches as described here (with a suitable relaxation of 
the priors on $\vT_{\rm cheb}$).  Figure \ref{fig:chebyshev} offers a practical demonstration of 
how these nuisance parameters are applied.

@@

The quality of the model spectrum is assessed by comparing to the data with a pixel-by-pixel 
likelihood calculation.  If we denote the data spectrum as $\vD$, then a corresponding residual 
spectrum (of $N_{\rm pix}$ pixels) can be defined for any input parameter set,
\begin{equation}
\vR \equiv \vR(\stellarParameters, \cheb) \equiv \vD-\vM(\stellarParameters, \cheb).
\end{equation}
To quantify the probability of the data conditioned on the model, we adopt a standard 
multi-dimensional Gaussian likelihood function,
\begin{equation}
p(\vD|\vM) =  \frac{1}{[(2 \pi)^{N_{\rm pix}} \det(\vC)]^{1/2}} \exp\left ( -\frac{1}{2} \,
   \vR^\trans \vC^{-1} \vR \right ), 
   \label{eqn:likelihood}
\end{equation}
which penalizes models that yield larger residuals and explicitly allows for covariances in the 
residual spectrum through the $N_{\rm pix} \times N_{\rm pix}$ matrix, $\vC$.  For practical 
numerical reasons, we generally consider the log-likelihood as the fit quality metric, which in 
this case is
\begin{equation}
  \ln{p(\vD | \vM)} = -\frac{1}{2} \left( \vR^\trans \vC^{-1} \vR + \ln{\det{\vC}} + N_{\rm pix} \ln{2\pi} \right).
  \label{eqn:lnlikelihood}
\end{equation}

The covariance matrix $\vC$ characterizes both the measurement uncertainty ($\sigma$; ``noise") in 
each pixel and the intrinsic covariance between pixels.  The special case where each pixel 
represents an independent measurement results in a diagonal covariance matrix, $\vC_{ij} = 
\delta_{ij} \,\, \sigma_i$ where $\sigma_i$ is the uncertainty in pixel $i$ and $\delta_{ij}$ is 
the Kronecker delta function, and Eq.~\ref{eqn:lnlikelihood} reduces to the familiar
\begin{equation}
\ln{p(\vD | \vM)} = -\frac{1}{2} \sum_i^{N_{\rm pix}} \frac{\vR_i^2}{\sigma_i^2} \equiv -\frac{\chi^2}{2},
\label{eqn:chisq}
\end{equation}
the sum of the square of the residuals weighted by the inverse variances (squared uncertainties).  
However, the problem being addressed here necessitates the use of a more complex covariance matrix; 
additional off-diagonal terms that can explicitly characterize (1) pixel-to-pixel covariances 
imposed by the discrete over-sampling of the line-spread function, and (2) highly correlated 
residuals as manifestations of the still-imperfect model library are required to avoid biasing our 
inferences of the physically interesting parameters ($\vt$).  The following subsections describe 
how these issues are addressed in the practical implementation of $\vC$.

>>>>

% Ian, this was ridiculous...don't type out giant words that are longer than the required tex.  It 
% actually takes a lot more time and is a huge pain.  
%\newcommand{\fundamentalParameters}{\vt_{\ast}}
%\newcommand{\gridParameters}{\{\fundamentalParameters\}^{\rm 
%grid}}
%\newcommand{\observationalParameters}{\vt_{\textrm{obs}}}
%\newcommand{\stellarParameters}{\vt}

%\phi and \Phi are used for nuisance parameters
\newcommand{\cheb}{ \vp_{\mathsf{P}}}
\newcommand{\chebi}[1]{ \vp_{\textrm{Cheb}_{#1}}} %for signifying a specific order
\newcommand{\Cheb}{ \vP_{\textrm{Cheb}}}
\newcommand{\Chebi}[1]{ \vP_{\textrm{Cheb}_{\ne #1}}} %for specifying everything but a specific order
\newcommand{\cov}{ \vp_{\mathsf{C}}}
\newcommand{\covi}[1]{ \vp_{\textrm{cov}_{#1}}} %for signifying a specific order
\newcommand{\Cov}{ \vP_{\textrm{cov}}}
\newcommand{\Covi}[1]{ \vP_{\textrm{cov}_{\ne #1}}} %for specifying everything but a specific order

@@

%kernel notation
\newcommand{\KK}{\mathcal{K}}
\newcommand{\Kglobal}{\KK_{\textrm{g}}}
\newcommand{\Klocal}{\KK_l}

@@

All astronomers recognize that spectroscopy offers a wealth of information that can help
characterize the properties of the observing target.  In the context of stellar astrophysics, 
spectroscopy plays many fundamental roles.  The relative strengths and widths of stellar absorption 
lines provide access to key parameters like effective temperature ($T_{\rm eff}$) and surface 
gravity ($\log g$), enabling model comparisons in the Hertzsprung-Russell diagram to estimate the 
masses and ages so crucial to understanding stellar evolution, as well as individual elemental (and 
molecular) abundances or the collective ``metallicity" (typically parameterized as [Fe/H]), 
facilitating study of the chemical hallmarks of different stellar populations.  With sufficient 
spectral resolution, the velocity content of a spectrum can convey crucial information about 
stellar rotation ($v \sin i$) and kinematics (e.g., association with a cluster or companion through 
the radial velocity, $v_z$).  While many fields benefit from the spectroscopic measurements of 
these stellar properties, there is acute interest in the exoplanet community.  There, all estimates 
of the planet properties are made {\it relative} to the host properties (e.g., the planet-to-host 
mass or radius {\it ratio} is constrained with the radial velocity or transit techniques, 
respectively).  Moreover, essential clues to the planet formation process are encapsulated in the 
dependences of planet frequency on host mass \citep[e.g.,][]{johnson07,howard10} and metallicity 
\citep[e.g.,][]{fischer05,buchhave14}.   
% SA: we ought to check with exoplanet people on these references. Can do later...

@@

The remainder of this section describes the mechanics of this modeling framework.  First, a model 
spectrum is generated for a given set of physical parameters (Section \ref{subsec:synthetic}), and 
then post-processed to mimic reality using a set of observational and practical nuisance parameters 
(Section \ref{subsec:postprocess}).  Next, a direct, pixel-by-pixel comparison between the data and 
model spectra is made with a prescribed likelihood function and a parametric treatment of the 
covariances between pixel residuals (Section \ref{subsec:likelihood}).  That process is iterated 
under the guise of hierarchical Markov Chain Monte Carlo (MCMC) simulations to numerically explore 
the posterior probability density of the model conditioned on the data, and thereby to determine 
constraints on the parameters of interest (Section \ref{subsec:MCMC}).  Along the way, these 
procedures are illustrated with real observations of the high resolution optical spectrum from a 
nearby F star.  That specific application, along with some alternative demonstrations of the 
method, are discussed in more detail in Section \ref{sec:examples}.

@@

There are various approaches to synthesizing a spectrum, $f_{\lambda}$, for a specific set of model
parameters, $\vt_{\ast}$.  In an ideal case, a model stellar atmosphere is constructed and then 
subsequently processed through a radiative transfer code \citep[e.g.,][]{kurucz93,hauschildt99}.  
However, in general this approach is still computationally prohibitive for any iterative method of 
probabilistic inference.  One partial compromise is to interpolate over a library of atmosphere 
structures that were pre-computed for a discrete grid of parameter values, $\{\vt_{\ast}\}^{\rm 
grid}$, for some arbitrary $\vt_{\ast}$, and then perform a radiative transfer calculation with 
that interpolated atmosphere to synthesize $\flam$ \citep[e.g., as for {\tt SME};][]{valenti96}.  A 
more common variant is to instead rely on interpolation over a library of pre-synthesized model 
spectra, $\flam(\vt_{\ast})$ \citep[e.g.,][]{castelli04,allard12,husser13}.  While technically the 
former approach is most similar to the ideal case, the computational cost of repeated spectral 
synthesis is sufficiently high to make a detailed exploration of parameter space (particularly for 
data with a large spectral range) considerably less appealing.  A related, but different approach 
is to eschew forward modeling entirely (and therefore repeated spectral syntheses and/or library 
interpolations), and instead evaluate the models only at the discrete grid points of the library.  
Then, these discretized samples of the posterior probability density can be interpolated to an 
arbitrary $\vt_{\ast}$ to construct appropriate confidence intervals \citep[similar to the method 
of {\tt SPC};][]{buchhave12}.  The difficulty with this latter approach is that the parameter 
uncertainties can be smaller than the grid spacing; in that case, there is valid concern that this 
interpolation might not accurately recover intrinsic parameter degeneracies.

Here we opt to take the computationally expedient approach that employs a library of model spectra, 
$\flam(\{\vt_{\ast}\}^{\rm grid})$, where $\vt_{\ast} = [T_{\rm eff}, \,\, \log{g}, \,\, Z]$ 
(in practice, the metallicity $Z$ is often parameterized by [Fe/H]).  However, it is worth noting 
that the techniques we will develop are applicable to {\it any} reasonable ``front-end" that 
generates a model spectrum.  In our adopted approach, the model spectrum for an arbitrary 
$\vt_{\ast}$ must be interpolated from among the spectral library, 
\begin{equation}
\flam(\{\vt_{\ast}\}^{\rm grid}) \leadsto \flam(\vt_{\ast}),
\label{eqn:interp}
\end{equation}
where we assign the symbol $\leadsto$ as an interpolation operator.  The multi-dimensional 
interpolation in Eq.~\ref{eqn:interp} needs to be performed many times, so computational efficiency 
is critical.  In practice, a simple tri-linear interpolation is suitably fast, but introduces an 
undesireable level of inaccuracy (particularly in the $Z$ dimension).  The interpolation quality 
can be empirically estimated by performing the operation in Eq.~\ref{eqn:interp} across a 
calculated location in $\{\vt_{\ast}\}^{\rm grid}$, and then comparing the interpolated spectrum 
with the corresponding library spectrum.  After an extensive exploration of such calculations 
\citep[see also][]{husser12}, we concluded that the best combination of speed and accuracy can be 
achieved by pre-computing a {\it refined} spectral library using a cubic spline interpolation with 
a $\{\vt_{\ast}\}^{\rm grid}$ spacing of [20\,K, 0.1\,dex, 0.1\,dex], and then performing 
tri-linear interpolation over that refined grid.  Overall, this interpolation technique is found to 
be accurate within a few percent per high resolution model pixel.  Ideally, this pre-interpolation 
could be avoided if the spectral library was computed over a refined grid (with a substantial 
up-front computational investment); but for the time being, we can empirically propagate these 
interpolation uncertainties into the likelihood calculations (as will be described in 
Section~\ref{subsec:likelihood}).

@@

Generally, the ``raw" model spectrum $\flam(\vt_{\ast})$ will be highly over-sampled compared to a 
typical observed spectrum, and does not account for several additional observational and 
instrumental effects that become important in comparisons with real data.  Therefore, a certain 
amount of post-processing is required before assessing the model quality.  We treat that 
post-processing in two stages: the first deals with an additional set of ``observational" 
parameters, $\vt_{\rm obs}$, that incorporate dynamical effects, geometry, and the relative 
location of the target, while the second employs a suite of nuisance (hyper-)parameters ($\vp$) 
designed to mitigate an imperfect data calibration.

We can further divide $\vt_{\rm obs}$ into those parameters that impact the model primarily in the 
spectral or flux dimensions; $\vt_{\rm obs} = [\vt_{\rm obs, v}, \,\, \vt_{\rm obs, f}]$.  For the 
former, we consider three kernels that contribute to the observed line-of-sight velocity 
distribution function, $\varphi_v$.  The first, $\mathcal{F}_v^{\rm inst}$, treats the instrumental 
spectral broadening.  For illustrative purposes we assume $\mathcal{F}_v^{\rm inst}$ is a Gaussian 
with a mean at $v = 0$ and a constant width $\sigma_v$ at all $\lambda$, although more 
sophisticated forms could be adopted.  The second, $\mathcal{F}_v^{\rm rot}$, characterizes the 
broadening induced by (projected) stellar rotation, parameterized by $v\sin{i}$ as described by 
\citet[][his Eq.~18.14]{gray08}.  And the third, $\mathcal{F}_v^{\rm dop} = \delta(v-v_z)$, 
incorporates the radial velocity through a Doppler shift.  The model spectrum is modified by the 
parameters $\vt_{{\rm obs}, v} = [\sigma_v, \,\, v\sin{i}, \,\, v_z]$ through these kernels, using 
a convolution in velocity-space,\footnote{In practice, these convolutions are performed as 
multiplications in Fourier-space to better preserve spectral information \citep[cf.,][]{tonry79}; 
the mathematical formalism is presented here for clarity.}
\begin{equation}
\flam(\vt_{\ast}, \vt_{{\rm obs}, v}) = \flam(\vt_{\ast}) \otimes \varphi_v = \flam(\vt_{\ast}) \otimes \mathcal{F}_v^{\rm inst} \otimes \mathcal{F}_v^{\rm rot} \otimes \mathcal{F}_v^{\rm dop},
\label{eqn:broadening}
\end{equation}
and then re-sampled onto the discrete wavelengths corresponding to each data pixel, 
\begin{equation}
\flam(\vt_{\ast}, \vt_{{\rm obs}, v}) \mapsto \vM(\vt_{\ast}, \vt_{{\rm obs}, v}),
\label{eqn:resampling}
\end{equation}
where the $\mapsto$ symbol denotes a re-sampling operator that maps the model spectrum onto the 
$N_{\rm pix}$-element model vector $\vM$ (and $N_{\rm pix}$ is the number of pixels in the 
spectrum).  For reference, Figure \ref{fig:broadening} shows a graphical representation of these 
post-processing steps.

@@

At this stage, the model is further modified in the flux dimension.  A typical synthetic spectrum 
is computed as the flux that would be measured {\it at the stellar surface}, and so needs to be 
diluted by the subtended solid angle, $\Omega = (R_{\ast}/d)^2$, where $R_{\ast}$ is the stellar 
radius and $d$ is the distance.  An additional wavelength-dependent scaling factor is applied to 
account for interstellar extinction, assuming some previously-derived extinction law $A_{\lambda}$ 
\citep[e.g.,][]{cardelli89} that is parameterized by $A_V$.  The parameters $\vt_{{\rm obs}, f} = 
[\Omega, \,\, A_V]$ are applied as
\begin{equation}
\vM(\vT) \equiv \vM(\vt_{\ast}, \vt_{{\rm obs},v}, \vt_{{\rm obs},f}) = \vM(\vt_{\ast}, \vt_{{\rm obs}, v}) \times \Omega \times 10^{-0.4\,A_{\lambda}},
\label{eqn:scaling}
\end{equation}
where we have simplified the notation by writing $\vT \equiv [\vt_{\ast}, \,\, \vt_{\rm obs}]$.

So far, the procedure summarized in Eq.~\ref{eqn:interp}-\ref{eqn:scaling} is composed of 
relatively straightforward operations demanded by practical astronomical and computing issues.  If 
the observed spectrum was {\it perfectly} calibrated, we could proceed to a likelihood calculation 
(Section~\ref{subsec:likelihood}) that makes a direct comparison with $\vM(\vT)$ at this point.  
However, that is unlikely to be the case.  The primary concern is that an imperfect calibration 
produces (presumably low-level) mismatches in the underlying shape of the observed spectrum on 
relatively broad wavelength scales.  When compared with the model, these mismatches might represent 
a non-trivial contribution to the likelihood, and thereby bias our estimates of the desired 
physical parameters.  Often, this concern is treated externally to any modeling procedure, usually 
by dividing the observed spectrum (and model spectrum) by an appropriate polynomial function.  But 
that ``normalization" procedure implicitly assumes that there is no relevant information content on 
those mismatched scales; if $\vT$ also contributes to the broad spectral shape, then adopting this 
approach will corrupt the inferences of these parameters.  Moreover, in practice this approach is 
limited, since defining an appropriate polynomial becomes difficult in cases where the spectral 
line density is high (e.g., molecular bands for cool stars).

We adopt a somewhat analogous approach to deal with this issue, but cast it {\it internal} to the 
modeling framework to appropriately propagate the uncertainty introduced by additional degrees of 
freedom in the model.  The residual calibration errors are treated as an explicit contribution to 
the model spectrum, enabling us to explore the distribution of possible calibrations (see 
Section~\ref{subsec:MCMC}) and then eventually marginalize out the associated nuisance parameters.  
In essence, the inferences for the relevant stellar parameters will properly account for the 
underlying uncertainty in the calibration process.  In practice, this is achieved by distorting 
segments of the model spectrum with polynomials, $\mathsf{P}$ \citep[e.g.,][]{eisenstein06,
koleva09}.  For data with $N_{\rm ord}$ spectral orders, each denoted with index $o$ (where $o$ 
naturally has an implicit $\lambda$-dependence), we can write
\begin{equation}
\vM(\vT, \cheb) = \vM(\vT) \times \{\mathsf{P}_o\} \equiv \vM(\vT) \times \left\{ \sum_n c_o^{(n)} \, T_o^{(n)} \right\},
\label{eqn:chebyshev}
\end{equation}
where $T^{(n)}$ are the standard Chebyshev functions of degree $n$ and the coefficients are treated 
as a set of $n \times N_{\rm ord}$ nuisance (hyper-)parameters, $\cheb = [\{\vp_{\mathsf{P}_o}\}] = 
[\{c_o^{(0)}, c_o^{(1)}, \ldots, c_o^{(n-1)} \}]$.  With judicious priors on $\cheb$, we can ensure 
that the unintended treatment of real spectral features (e.g., broad, deep molecular bands) as 
calibration artifacts is negligible (see Section~\ref{sec:examples} for examples).  The 
lowest-degree (scaling) coefficient, $c^{(0)}$, is by its nature degenerate with the solid angle 
parameter, $\Omega$.  Therefore, we enforce an additional constraint by requiring that the mean of 
the polynomial is unity.  For data with a single spectral order, this means simply setting $c^{(0)} 
= 1$.  If the goal is to model multiple spectral orders, we assign $c^{(0)} = 1$ in an arbitrary 
order as an anchor, but permit the $c^{(0)}$ values in other orders to be different (as 
necessary).  It is worth noting that this formalism can, in principle, be extended to develop 
models for completely uncalibrated spectra, rather than the residual calibration mismatches as 
described here (with a suitable relaxation of the priors on $\cheb$).  Figure \ref{fig:chebyshev} 
offers a practical demonstration of how these nuisance parameters are applied.

@@

The quality of the model spectrum is assessed by comparing to the data with a pixel-by-pixel 
likelihood calculation.  If we denote the data spectrum as $\vD$, then a corresponding residual 
spectrum (vector of $N_{\rm pix}$ elements) can be defined for any input parameter set,
\begin{equation}
\vR \equiv \vR(\vT, \cheb) \equiv \vD-\vM(\vT, \cheb).
\end{equation}
To quantify the probability of the data conditioned on the model, we adopt a standard 
multi-dimensional Gaussian likelihood function,
\begin{equation}
p(\vD|\vM) =  \frac{1}{[(2 \pi)^{N_{\rm pix}} \det(\vC)]^{1/2}} \exp\left ( -\frac{1}{2} \,
   \vR^\trans \vC^{-1} \vR \right ), 
   \label{eqn:likelihood}
\end{equation}
which penalizes models that yield larger residuals and explicitly allows for covariances in the 
residual spectrum through the $N_{\rm pix} \times N_{\rm pix}$ matrix, $\vC$.  For practical 
numerical reasons, we generally consider the log-likelihood as the fit quality metric, which in 
this case is
\begin{equation}
  \ln{p(\vD | \vM)} = -\frac{1}{2} \left( \vR^\trans \vC^{-1} \vR + \ln{\det{\vC}} + N_{\rm pix} \ln{2\pi} \right).
  \label{eqn:lnlikelihood}
\end{equation}

The covariance matrix $\vC$ characterizes both the measurement uncertainty ($\sigma$; ``noise") in 
each pixel and the intrinsic covariance between pixels.  The special case where each pixel 
represents an independent measurement results in a diagonal covariance matrix, $\vC_{ij} = 
\delta_{ij} \,\, \sigma_i$ where $\sigma_i$ is the uncertainty in pixel $i$ and $\delta_{ij}$ is 
the Kronecker delta function, and Eq.~\ref{eqn:lnlikelihood} reduces to the familiar
\begin{equation}
\ln{p(\vD | \vM)} = -\frac{1}{2} \sum_i^{N_{\rm pix}} \frac{\vR_i^2}{\sigma_i^2} \equiv -\frac{\chi^2}{2},
\label{eqn:chisq}
\end{equation}
the sum of the square of the residuals weighted by the inverse variances (squared uncertainties).  
However, the problem being addressed here necessitates the use of a more complex covariance matrix; 
additional off-diagonal terms that can explicitly characterize (1) pixel-to-pixel covariances 
imposed by the discrete over-sampling of the line-spread function, and (2) highly correlated 
residuals as manifestations of the still-imperfect model library are required to avoid biasing our 
inferences of the physically interesting parameters ($\vT$).  The following subsections describe 
how these issues are addressed in the practical implementation of $\vC$.
